### 2014-10-23 Paper_v1.tex
Near final draft

\section*{Introduction}
Developing mathematical models of biochemical networks is a significant facet of systems biology.
Modeling approaches differ in their degree of detail, where the choice of approach is often determined by prior knowledge, or model requirements \citep{2012_kholodenko_kolch_SciSig}. 
Ordinary differential equation (ODE) models are a common tool for modeling biochemical systems because of their ability to capture dynamics, and encode mechanism.
However, ODE models typically come with difficult (or sometimes impossible) parameter identification problems. 
For example, Gadkar et al., showed that even with near-perfect information, 
it was often impossible to identify all the parameters in typical signal transduction models \citep{Gadkar:2005ad}. 
However, it is not clear whether we actually need precise estimates for all model parameters. 
Bailey suggested more than a decade ago, that achieving qualitative or even quantitative understanding of biological systems 
should not require complete structural and parametric knowledge \citep{2001_bailey_NatBiotech}. 
Since Bailey's complex biology with no parameters hypothesis, Sethna showed that model performance is typically 
sensitive to only a few parameters, a characteristic seemingly universal to multi-parameter models referred to as \textit{sloppiness} \citep{Machta:2013aa}.
Thus, reasonable predictions may be possible, despite parametric uncertainty, if a few critical parameters are well defined. 
Perhaps more importantly, ODE models also require significant mechanistic information, thereby limiting their utility in poorly understood systems.
Toward this challenge, logical modeling is an emerging paradigm that encodes causal relationships using non-linear transfer functions \citep{Morris:2010aa}.
These functions encode interactions between model components, without requiring mechanism.
Despite their simplicity, logical models have captured rich system behaviors in the absence of mechanistic information \citep{Saez-Rodriguez:2011aa,Morris:2011ys,Morris:2012aa}. 
However, modeling complex dynamics remains difficult with this class of model.
Thus, there is an unmet need for a third approach which combines ODEs and logical models, where ODEs could encode mechanistic information, 
while missing or incomplete mechanistic knowledge can be approximated using a logical approach.

In this study, we developed a hybrid approach which combines ODE modeling with logical rules to model a well studied biochemical network, 
the human coagulation system. Coagulation is an archetype proteolytic cascade involving both positive 
and negative feedback \cite{Butenas:2002aa,Schenone:2004aa,Adams:2009aa}. 
Coagulation is mediated by a family proteases in the circulation, called factors and a key group of blood cells, called platelets. 
The central process in coagulation is the conversion of prothrombin (fII), an inactive coagulation factor, to the master protease thrombin (FIIa).
Thrombin generation involves three phases, initiation, amplification and termination \cite{GOLDHABER2006, Brummel:2002aa}.
Initiation requires a trigger event, for example vessel injury, which leads to the activation of factor VII (FVIIa).
Two converging pathways, the extrinsic and intrinsic cascades, then process and amplify this initial coagulation signal. 
The extrinsic cascade is generally believed to be the main mechanism of thrombinogenesis in the blood \cite{MANN1990,ROBERTS1998,MANN1999}.
Initially, thrombin is produced upon cleavage of prothrombin by fluid phase activated factor X (FXa), which itself has been activated by FVIIa \cite{Butenas:2002aa}. 
Picomolar amounts of thrombin then activate the cofactors factors V and VIII (fV and fVIII) and platelets, 
leading to the formation of the tenase and prothrombinase complexes on activated platelets.
These complexes amplify the early coagulation signal by further activating FXa, and directly converting prothrombin to thrombin. 
There are several control points in the cascade that inhibit thrombin formation, and eventually terminate thrombin generation. 
Tissue Factor Pathway Inhibitor (TFPI) inhibits FXa formation catalyzed by TF/FVIIa, while antithrombin III (ATIII)
neutralizes several of the proteases generated during coagulation, including thrombin.  
Thrombin itself also inadvertently plays a role in its own inhibition; thrombin, through interaction with thrombomodulin, protein C and endothelial cell protein C receptor (EPCR),
converts protein C to activated protein C (APC) which attenuates the coagulation response by proteolytic cleavage of fV/FVa and fVIII/FVIIIa. 
Termination occurs after either prothrombin is consumed, or thrombin formation is neutralized by inhibitors such as activated protein C (APC) or antithrombin III (ATIII).

@@

\subsection*{Formulation of reduced order coagulation models.}
We developed a reduced order coagulation model to test our hybrid kinetic rule based modeling approach (Fig. \ref{fig-network}).
A trigger event initiates thrombin formation (FIIa) from prothrombin (fII) through a lumped initiation step.
This step loosely represents the initial fluid phase activation of thrombin by activated FXa.
Once activated, thrombin catalyzes its own formation (amplification step), and inhibition via the conversion of protein C to activated protein C (APC).
APC and tissue factor pathway inhibitor (TFPI) inhibit initiation, while antithrombin III (ATIII) inhibits amplification. 
All initiation and inhibition processes, as well as the dependence of amplification upon other coagulation factors, was approximated using our rule-based approach (Fig. \ref{fig-control-schematic}). 
Individual regulatory contributions to the activity of pathway enzymes were integrated into control coefficients using an integration rule.  
These control coefficients then modified the rates of model processes at each time step.
Hill-like transfer functions $0 \leq f\left(\mathcal{Z}\right) \leq 1$ quantified the influence of components upon a target process. 
Components were either individual inhibitor or activator levels or some function of levels, e.g., the product of factor levels. 
In this study, $\mathcal{Z}$ corresponded to the abundance of individual inhibitors or activators, 
with the exception of the dependence of amplification upon specific coagulation factors (modeled as the product of factors). 
When a process was potentially sensitive to multiple inputs, logical integration rules were used to select which transfer functions influenced the process at any given time. 
In our proof of concept model, we used a winner takes all strategy; the maximum or minimum transfer function was selected at any given time step. 
However, other integration rules are certainly possible. 
Taken together, while the reduced order coagulation model encodes significant biological complexity, it is highly compact, consisting of only five differential equations. 
Thus, it will serve as an excellent proof of principle example to study the reduction of a highly complex human subsystem.

@@

\subsection*{Estimation of model parameters from experimental data.}
Model parameters were estimated by minimizing the difference between simulations and experimental thrombin measurements (squared residual):
\begin{equation}\label{eqn:objective-function}
	\min_{\mathbf{k}} \sum_{\tau=1}^{\mathcal{T}}\sum_{j=1}^{\mathcal{S}}\left(\frac{\hat{x}_{j}\left(\tau\right) - x_{j}\left(\tau,\mathbf{k}\right)}{\omega_{j}\left(\tau\right)}\right)^{2}
\end{equation}where $\hat{x}_{j}\left(\tau\right)$ denotes the measured value of species $j$ at time $\tau$, $x_{j}\left(\tau,\mathbf{k}\right)$ denotes the simulated 
value for species $j$ at time $\tau$, and $\omega_{j}\left(\tau\right)$ denotes the experimental measurement variance for species $j$ at time $\tau$. The outer summation is respect to
time, while the inner summation is with respect to state. We minimized the model residual using Particle swarm optimization (PSO) \citep{PSO}.
PSO uses a \textit{swarming} metaheuristic to explore parameter spaces. 
A strength of PSO is its ability to find the global minimum, even in the presence of potentially many local minima, by communicating the local
error landscape experienced by each particle collectively to the swarm. Thus, PSO acts both as a local and a global search algorithm. 
For each iteration, particles in the swarm compute their local error by evaluating the model equations using their specific parameter vector realization.
From each of these local points, a globally best error is identified. Both the local and global error 
are then used to update the parameter estimates of each particle using the rules:
\begin{eqnarray}
	\mathbf{\Delta}_{i} &=&\theta_{1}\mathbf{\Delta}_{i} + \theta_{2}\mathbf{r}_{1}\left(\mathcal{L}_{i} - \mathbf{k}_{i}\right) + \theta_{3}\mathbf{r}_{2}\left(\mathcal{G} - \mathbf{k}_{i}\right) \\
	\mathbf{k}_{i} &=& \mathbf{k}_{i} + \mathbf{\Delta}_{i}
\end{eqnarray}where $\left(\theta_{1},\theta_{2},\theta_{3}\right)$ are adjustable parameters, $\mathcal{L}_{i}$ denotes local best solution found by particle $i$, and
$\mathcal{G}$ denotes the best solution found over the entire population of particles. The quantities $r_{1}$ and $r_{2}$ denote uniform random vectors with the same dimension as the number of unknown model
parameters ($\mathcal{K}\times{1}$). In thus study, we used $\left(\theta_{1},\theta_{2},\theta_{3}\right) = \left(1.0, 0.05564, 0.02886\right)$. The quality of parameter
estimates was measured using goodness of fit (model residual). The particle swarm optimization routine was implemented in the Python programming language. 
All plots were made using the Matplotlib module of Python \citep{Matplotlib}.
The optimization code can be downloaded from http://www.varnerlab.org.

\subsection*{Global sensitivity analysis of model performance.}
We conducted a global sensitivity analysis to estimate which parameters controlled the performance of the reduced order model. 
We computed the total sensitivity index of each parameter relative to two performance objectives, the peak thrombin time and the area under the thrombin curve 
(thrombin exposure). [FINISH~ME].

@@

\begin{figure}
\centering
\includegraphics[width=1.0\textwidth]{./figs/Figure-7-rFVIIa-Trigger.pdf}
\caption{Reduced order coagulation model predictions of rFVIIa administration. 
A: Simulations of thrombin formation in the presence of ATIII and the protein C pathway were conducted for a range of trigger values (1x - 200x nominal) in the absence of fVIII and fIX.
B: Comparison of thrombin generation for normal versus hemophilia for 10x nominal trigger.
Solid lines denote the simulated mean value of the thrombin profile for N = 20 independent particles.
The shaded region denotes the 99\% confidence estimate of the mean simulated thrombin value (uncertainty in the model simulation). 
All factors and control proteins were at their physiological concentration unless others noted.}\label{fig-rFVII}
\end{figure}

>>>>

\section*{Introduction}
Developing mathematical models of biochemical networks is a significant facet of systems biology.
Modeling approaches differ in their degree of detail, where the choice of approach is often determined by prior knowledge, or model requirements \citep{2012_kholodenko_kolch_SciSig}. 
Ordinary differential equation (ODE) models are a common tool for modeling biochemical systems because of their ability to capture dynamics, and encode mechanism.
However, ODE models typically come with difficult (or sometimes impossible) parameter identification problems. 
For example, Gadkar et al., showed that even with near-perfect information, 
it was often impossible to identify all the parameters in typical signal transduction models \citep{Gadkar:2005ad}. 
However, it is not clear whether we actually need precise estimates for all model parameters. 
Bailey suggested more than a decade ago, that achieving qualitative or even quantitative understanding of biological systems 
should not require complete structural and parametric knowledge \citep{2001_bailey_NatBiotech}. 
Since Bailey's complex biology with no parameters hypothesis, Sethna showed that model performance is typically 
sensitive to only a few parameters, a characteristic seemingly universal to multi-parameter models referred to as \textit{sloppiness} \citep{Machta:2013aa}.
Thus, reasonable predictions may be possible, despite parametric uncertainty, if a few critical parameters are well defined. 
Perhaps more importantly, ODE models also require significant mechanistic information, thereby limiting their utility in poorly understood systems.
Toward this challenge, logical modeling is an emerging paradigm that encodes causal relationships between model components using non-linear transfer functions, 
without requiring mechanism. \citep{Morris:2010aa}. 
Despite their simplicity, logical models have captured rich behaviors in a variety of systems important to human health \citep{Saez-Rodriguez:2011aa,Morris:2011ys,Morris:2012aa}. 
However, modeling complex dynamics with logical models is challenging. 
Thus, there is an unmet need for a third approach which combines ODEs and logical models, where ODEs could encode mechanistic information, 
while missing or incomplete mechanistic knowledge can be approximated using a logical approach.

In this study, we developed a hybrid approach which combines ODE modeling with logical rules to model a well studied biochemical network, 
the human coagulation system. Coagulation is an archetype proteolytic cascade involving both positive 
and negative feedback \cite{Butenas:2002aa,Schenone:2004aa,Adams:2009aa}. 
Coagulation is mediated by a family proteases in the circulation, called factors and a key group of blood cells, called platelets. 
The central process in coagulation is the conversion of prothrombin (fII), an inactive coagulation factor, to the master protease thrombin (FIIa).
Thrombin generation involves three phases, initiation, amplification and termination \cite{GOLDHABER2006, Brummel:2002aa}.
Initiation requires a trigger event, for example vessel injury, which leads to the activation of factor VII (FVIIa).
Two converging pathways, the extrinsic and intrinsic cascades, then process and amplify this initial coagulation signal. 
The extrinsic cascade is generally believed to be the main mechanism of thrombinogenesis in the blood \cite{MANN1990,ROBERTS1998,MANN1999}.
Initially, thrombin is produced upon cleavage of prothrombin by fluid phase activated factor X (FXa), which itself has been activated by TF/FVIIa \cite{Butenas:2002aa}. 
Picomolar amounts of thrombin then activate the cofactors factors V and VIII (fV and fVIII) and platelets, 
leading to the formation of the tenase and prothrombinase complexes on activated platelets.
These complexes amplify the early coagulation signal by further activating FXa, and directly converting prothrombin to thrombin. 
There are several control points in the cascade that inhibit thrombin formation, and eventually terminate thrombin generation. 
Tissue Factor Pathway Inhibitor (TFPI) inhibits FXa formation catalyzed by TF/FVIIa, while antithrombin III (ATIII)
neutralizes several of the proteases generated during coagulation, including thrombin.  
Thrombin itself also inadvertently plays a role in its own inhibition; thrombin, through interaction with thrombomodulin, protein C and endothelial cell protein C receptor (EPCR),
converts protein C to activated protein C (APC) which attenuates the coagulation response by proteolytic cleavage of fV/FVa and fVIII/FVIIIa. 
Termination occurs after either prothrombin is consumed, or thrombin formation is neutralized by inhibitors such as activated protein C (APC) or antithrombin III (ATIII).

@@

\subsection*{Formulation of reduced order coagulation models.}
We developed a reduced order extrinsic coagulation model to test our hybrid modeling approach (Fig. \ref{fig-network}).
A trigger event initiates thrombin formation (FIIa) from prothrombin (fII) through a lumped initiation step.
This step loosely represents the initial activation of thrombin by activated FXa.
Once activated, thrombin catalyzes its own formation (amplification step), and inhibition via the conversion of protein C to activated protein C (APC).
Antithrombin III (ATIII) inhibits amplification, while APC and tissue factor pathway inhibitor (TFPI) potentially inhibit both initiation and amplification. 
All initiation and inhibition processes, as well as the dependence of amplification upon other coagulation factors, was approximated using our rule-based approach (Fig. \ref{fig-control-schematic}). 
Individual regulatory contributions to the activity of pathway enzymes were integrated into control coefficients (v's) using an integration rule (min/max).  
These control coefficients then modified the rates of model processes at each time step.
Hill-like transfer functions $0 \leq f\left(\mathcal{Z}\right) \leq 1$ quantified the contribution of components upon a target process. 
Components were either individual inhibitor or activator levels or some function of levels, e.g., the product of factor levels. 
In this study, $\mathcal{Z}$ corresponded to the abundance of individual inhibitors or activators, 
with the exception of the dependence of amplification upon specific coagulation factors (modeled as the product of factors). 
When a process was potentially sensitive to multiple inputs, logical integration rules were used to select which transfer functions influenced the process at any given time. 
In our proof of concept model, we used a winner takes all strategy; the maximum or minimum transfer function was selected at any given time step. 
However, other integration rules are certainly possible. 
Taken together, while the reduced order coagulation model encodes significant biological complexity, it is highly compact (consisting of only five differential equations). 
Thus, it will serve as an excellent proof of principle example to study the reduction of a highly complex human subsystem.

@@

\subsection*{Estimation of model parameters from experimental data.}
Model parameters were estimated by minimizing the difference between simulations and experimental thrombin measurements (squared residual):
\begin{equation}\label{eqn:objective-function}
	\min_{\mathbf{k}} \sum_{\tau=1}^{\mathcal{T}}\sum_{j=1}^{\mathcal{S}}\left(\frac{\hat{x}_{j}\left(\tau\right) - x_{j}\left(\tau,\mathbf{k}\right)}{\omega_{j}\left(\tau\right)}\right)^{2}
\end{equation}where $\hat{x}_{j}\left(\tau\right)$ denotes the measured value of species $j$ at time $\tau$, $x_{j}\left(\tau,\mathbf{k}\right)$ denotes the simulated 
value for species $j$ at time $\tau$, and $\omega_{j}\left(\tau\right)$ denotes the experimental measurement variance for species $j$ at time $\tau$. The outer summation is respect to
time, while the inner summation is with respect to state. We minimized the model residual using Particle swarm optimization (PSO) \citep{PSO}.
PSO uses a \textit{swarming} metaheuristic to explore parameter spaces. 
A strength of PSO is its ability to find the global minimum, even in the presence of potentially many local minima, by communicating the local
error landscape experienced by each particle collectively to the swarm. Thus, PSO acts both as a local and a global search algorithm. 
For each iteration, particles in the swarm compute their local error by evaluating the model equations using their specific parameter vector realization.
From each of these local points, a globally best error is identified. Both the local and global error 
are then used to update the parameter estimates of each particle using the rules:
\begin{eqnarray}
	\mathbf{\Delta}_{i} &=&\theta_{1}\mathbf{\Delta}_{i} + \theta_{2}\mathbf{r}_{1}\left(\mathcal{L}_{i} - \mathbf{k}_{i}\right) + \theta_{3}\mathbf{r}_{2}\left(\mathcal{G} - \mathbf{k}_{i}\right) \\
	\mathbf{k}_{i} &=& \mathbf{k}_{i} + \mathbf{\Delta}_{i}
\end{eqnarray}where $\left(\theta_{1},\theta_{2},\theta_{3}\right)$ are adjustable parameters, $\mathcal{L}_{i}$ denotes local best solution found by particle $i$, and
$\mathcal{G}$ denotes the best solution found over the entire population of particles. The quantities $r_{1}$ and $r_{2}$ denote uniform random vectors with the same dimension as the number of unknown model
parameters ($\mathcal{K}\times{1}$). In thus study, we used $\left(\theta_{1},\theta_{2},\theta_{3}\right) = \left(1.0, 0.05564, 0.02886\right)$. The quality of parameter
estimates was measured using goodness of fit (model residual). The particle swarm optimization routine was implemented in the Python programming language. 
All plots were made using the Matplotlib module of Python \citep{Matplotlib}.
%The optimization and plotting code can be downloaded from http://www.varnerlab.org.

\subsection*{Global sensitivity analysis of model performance.}
We conducted a global sensitivity analysis, using the variance based method of Sobol, to estimate which parameters controlled the performance of the reduced order model \citep{SOBOL_METHOD}. 
We computed the total sensitivity index of each parameter relative to two performance objectives, the peak thrombin time and the area under the thrombin curve (thrombin exposure).
We established the sampling bounds for each parameter from the minimum and maximum value of that parameter in the parameter set ensemble. 
We used the sampling method of Saltelli et al., \citep{Saltelli:2010} to compute a family of $N\left(2d+2\right)$ parameter sets which obeyed our parameter ranges, 
where $N$ was the number of trials, and $d$ was the number of parameters in the model. In our case, N = 10,000 and d = 22, so the total sensitivity indices were computed from
460,000 model evaluations. The variance based sensitivity analysis was conducted using the SALib module encoded the Python programming language \citep{SALIB}.

@@

\begin{figure}
\centering
\includegraphics[width=1.0\textwidth]{./figs/Figure-7-rFVIIa-Trigger.pdf}
\caption{Reduced order coagulation model predictions of rFVIIa administration. 
A: Simulations of thrombin formation in the presence of ATIII and the protein C pathway were conducted for a range of trigger values (1x - 200x nominal) in the absence of fVIII and fIX.
B: Comparison of thrombin generation for normal versus hemophilia for 10x nominal trigger.
Solid lines denote the simulated mean value of the thrombin profile for N = 20 independent particles.
The shaded region denotes the 99\% confidence estimate of the mean thrombin value (uncertainty in the model simulation). 
All factors and control proteins were at their physiological concentration unless others noted.}\label{fig-rFVII}
\end{figure}
